Predicting detection probabilities for rare and undersampled North American landbirds

### 2023-12-07 article/article.tex
updated methods, results, and discussion of article based on j and a suggestions. still waiting for distance CV to finish

\par To test prediction \ref{pred1}, we ran the Bayesian hierarchical QPAD Removal and Distance models on the set of species for which there was sufficient data in the original NA-POPS analysis.
In this case, we ran both models for 316 species of birds.
Rufous Hummingbird (\textit{Selasphorus rufus}) was dropped from the distance modelling due to issues with divergent transitions \citep{betancourt_diagnosing_2016, leimkuhler_simulating_2005}.
Tricoloured Blackbird had sufficient distance data, and so was included in the distance modelling, but not the removal modelling.
We accessed the single-species estimates of cue rate and EDR from the NA-POPS Null models using the \texttt{napops} R package \citep{edwards_napops_2023, edwards_point_2023}.
We then modelled the difference in estimated cue rates and EDRs as
$$\Delta X \sim N(0,\sigma)$$
$$\sigma \sim exp(1)$$

where $\Delta X$ is calculated as the mean multi-species estimate of cue rate or EDR minus the mean single-species estimate of cue rate of EDR.
If prediction \ref{pred1} is true, we would expect to see a distribution of modelled differences centred about 0, indicating no real difference in estimates of cue rate or EDR between single-species and multi-species models, for species for which we have sufficient data.

\par To test prediction \ref{pred2}, we modelled the standard deviations about each estimate of cue rate or EDR from both the multi-species and single-species models, and regressed them against the log sample size per species.
That is
$$SD \sim N(\alpha_M + \beta_M\times \log(N), \sigma)$$
$$\alpha, \beta \sim N(0,1)$$
$$\sigma \sim exp(1)$$

where $\alpha$ is the model-specific intercept and $\beta$ is the model-specific slope.
In both cases, $M=1$ corresponds to the single species model and $M=2$ corresponds to the multi-species model.

\par If prediction \ref{pred2} is true, we would expect to see a higher value of $\alpha_1$ than for $\alpha_2$, corresponding to a higher baseline standard deviation in the single species models.
We would also likely see a slightly smaller (i.e., more negative) value of $\beta_1$ than $\beta_2$ (which should also be negative), indicating that the standard deviation decreases with log sample size for both models, but should decrease at a slower rate with the multi-species model (which has a lower baseline standard deviation to begin with) than the single-species model.

\par To test prediction \ref{pred3}, we ran a k-fold cross-validation with 5 folds, stratified by species. That is, for each cross-validation run, we held out 20\% of the data for each species for both the multi-species and Bayesian single-species models. 
We then derived multi-species estimates of cue rate and EDR and single-species estimates of cue rate and EDR for each cross-validation run, and used these parameter estimates to calculate the posterior predictive density of each of the held out data points. 
We then calculated the difference in posterior predictive density for removal and distance models, by subtracting the pointwise posterior predictive density of the single-species model from the pointwise posterior predictive density of the multi-species model (i.e., multi-species minus single-species).
Finally, we took the mean of the difference in pointwise posterior predictive density for each species to create a species-specific mean difference in pointwise predictive density.
With this, positive values would indicate that the multi-species model was preferred, and negative values would indicate that the single-species model was preferred.
If prediction \ref{pred3} is true, then we would expect to see majority positive values of the species-specific mean pointwise predictive density.

\subsection{Predicting Detection Probabilities}
\par To predict estimates of cue rate or EDR from the multi-species models, we adopted two possible approaches.
If the NA-POPS database contained some point-count data for the species of interest, we simply added these data into the model.
This is different from \citet{edwards_point_2023, solymos_calibrating_2013, solymos_evaluating_2018}, where any species with less than 75 data points were removed from the analysis.
For species that did not have any data in the NA-POPS database, we generated a row of traits data for the species and estimated its cue rate and EDR as parameters in the model.
We then ran the multi-species models described above using these additional data points (and hence an updated phylogenetic correlation structure for the removal model) to generate estimates of cue rate and EDR for these new species.

\subsection{Analysis}
\par All analyses were performed in R version 4.2.2 \citep{r_core_team_r_2022}.
Manipulation of phylogenetic trees, including finding a consensus tree for visualization and generating variance-covariance matrices based on the 2000 bootstrapped trees, was done using the \texttt{ape} R package \citep{paradis_ape_2019}.
All Bayesian models were coded in the Stan Probabilistic Language \cite{stan_development_team_stan_2019}, and models were run using the \texttt{cmdstanr} R package \citep{gabry_cmdstanr_2022} and visualized using the \texttt{bayesplot} R package \citep{gabry_visualization_2019}.
Each of the single-species models and multi-species models running both the initial comparison set of species and the final predictive set of species, as well as the models created to assess performance, were run using 1000 warmup iterations and 2000 sampling iterations, each on 4 chains for a total of 8000 draws per model run.
To mitigate issues surrounding initial values being rejected in the distance model (see Discussion), we set the initial values of EDR to be the mean EDR from NA-POPS estimates; for species where EDR estimates did not exist in NA-POPS, we used the overall mean EDR as initial value.
Cross-validation models were run with 1000 warmup iterations and 500 sampling iterations, each on 4 chains for a total of 2000 draws per model run.
All of the models made use of Stan's reduce\_sum functionality, which allows for within-chain parallelization as well as the usual across-chain parallelization.
This was particularly useful in our case as the multi-species removal model contained over 3 million data points and the multi-species distance model contained over 4 million data points, and so computational expense was a large concern.
All models were run on one of two multi-core processing servers, one running Ubuntu 20.04.4 LTS and one running Ubuntu 20.04.6 LTS.

@@

\par (NEEDS UPDATING) We found some evidence that a multi-species removal model improved the precision of cue rate estimates compared to estimates from a single-species model.
The intercept of standard deviations of cue rates estimated in a single species model was greater than the intercept of standard deviations of cue rates estimated in a multi-species model, with very little overlap in their 95\% credible intervals (Figure \ref{fig:sd}A).
This implies that the baseline standard deviation of cue rates is higher in a single species model than a multi species model for very low sample sizes.
However, the slope of standard deviations of cue rates estimated in a single species model was farther from 0 (i.e., more negative) than the slope of standard deviations of cue rates estimated in a multi-species model, with very little overlap in their 95\% credible intervals (Figure \ref{fig:sd}A).
This implies that the decrease in standard deviation as sample size increases is lesser in a multi-species model than a single-species model.
This indicates that after a certain sample size, it is likely that the difference in standard deviation between a single species and multi species model would be negligible.

\par (NEEDS UPDATING) We found little evidence that a multi-species model improves the precision of EDR compared to that of a single-species model.
The 95\% credible intervals for both estimated slopes and intercepts overlapped with each other, indicating that the standard deviations are effectively the same for a single-species and multi-species estimate as sample size increases (Figure \ref{fig:sd}A).

@@

\par The cross validation revealed that single species model actually had better predictive capacity for more species (222 out of 316) than the multi-species model (94 out of 316), but the difference in predictive capacity was overall very small.
Mean overall difference in posterior predictive density was $-2.21 \times 10^{-5}$.
Single-species models tended to have better predictive capacity when sample size was low (again noting a very small difference), and a negligible difference as sample size increased (Figure \ref{fig:cv}A).
There did not appear to be any patterns in model selection among phylogeny or traits (see Supplemental Figure 1 for full phylogenetic tree).
Six out of the 7 species that had a percent change in cue rate of more than 20\% had the multi-species model chosen as the better predictive model.
Common Redpoll was the only species that had the single-species model chosen (but please see Discussion about this particular species).

\par The cross validation revealed that the multi-species model tended to have slightly more support than the single-species model in terms of predicting EDR.
However, similar to the removal model cross-validation results, the overall difference in posterior predictive density was very small, with a mean difference of $1.85 \times 10^{-4}$.
Multi-species models tended to have better predictive capacity when sample size was low (again noting a very small difference), and a negligible difference as sample size increased (Figure \ref{fig:cv}B).
Interestingly, only two (Chimney Swift and Boat-tailed Grackle) of the six species with a greater than 20\% change in EDR had the multi-species model chosen as the better predictive model.

@@

\par The multi-species removal model produced reasonable estimates of cue rates for all seven species of bird considered here (Figure \ref{fig:predictions}A).
Species where the sample size requirements were relaxed (i.e., all except Bicknell's Thrush and Kirtland's Warbler) had reasonable precision associated with the estimates.
One exception to this was LeConte's Thrasher, where the precision was rather low.
The species for which we had no data (i.e., Bicknell's Thrush and Kirtland's Warbler) had estimates of cue rate similar to that of their closely-related species, demonstrating the shrinkage effect of the multi-species model.

\par The multi-species distance model also produced reasonable estimates of EDR for all six species of landbird considered here (Figure \ref{fig:predictions}B).
Species where the sample size requirements were relaxed (i.e., all except Bicknell's Thrush) had fairly precise estimates, except for LeConte's Thrasher and Lesser Prairie Chicken where the estimated EDR spanned greater than 50 m in either direction.
The species for which we had no data (i.e., Bicknell's Thrush) had an estimate of EDR similar to what would be expected by a bird of its size, song pitch, and habitat association, demonstrating the shrinkage effect of the multi-species model.

@@

@@

\par The multi-species QPAD Distance model had several issues related to initialization, sampling, and convergence throughout this study.
For many iterations of the model development, Stan's generation of initial values proved difficult.
For all unconstrained parameters, Stan draws initial values from a uniform(-2,2) distribution.
On the scale that we modelled detection distances (i.e., log detection distance in hundreds of metres), these are reasonable values.
At the low end, an initial value of -2 would correspond to a detection distance of $e^{-2} \times 100 = 13.5$ m, which is low but somewhat plausible for a small-bodied, high-pitched species such as a hummingbird.
At the high end, an initial value of 2 would correspond to a detection distance of $e^{2} \times 100 = 739$ m, which is high but still somewhat plausible for large-bodied, low-pitched species like a grouse or prairie-chickens.
However, where Stan seemed to struggle is when these randomly assigned values for detection distances were effectively impossible for a large set of species.
That is to say, if the values were initialized such that hummingbirds had a detection distance of 500 m or more, and grouse or prairie-chickens had a detection distance of 15 m or less, this would result in an impossible likelihood, forcing Stan to reject the initial values.
Our fix was to set initial values to be jittered mean detection distances originally estimated in \citet{edwards_point_2023}; for species without estimates, we used the mean EDR for all species.
This approach seemed to work as the initial values were no longer rejected, but we caution that this could very well be a "bandaid solution" to an underlying issue with the model itself.

\par The distance models also seemed to struggle with sampling the posterior distribution, in that over a quarter of the iterations resulted in a divergent transition \citep{betancourt_diagnosing_2016, leimkuhler_simulating_2005}.
This is likely due to hierarchical nature of the model and the parameterization choices that we made.
\citet{betancourt_hierarchical_2020} discuss at length some issues that arise when sampling from hierarchical models in cases where the hierarchical groups are data rich, where a "centred parameterization" should be used, or data poor, where a "non-centred parameterization" should be used.
We experimented with using only a non-centred parameterization which resulted in extremely low effective sample sizes (as low as 4 for 8000 iterations), large R-hat values, and took an unreasonable amount of time to finish sampling (approximately 2.5 weeks).
The model presented in this paper is a centred parameterization, which produced much more reasonable effective sample sizes and R hat values, finished within a day and a half, but still suffered from degeneracies such as hierarchical funnelling as described by \citet{betancourt_hierarchical_2020}.
In our case, because we are modelling on a hierarchy that contain many data points for some species x trait groups and fewer data points for other species x trait groups, our divergent transitions may be resolved if we were to implement our model using a mixture of centred and non-centred parametrization as described by \citet{betancourt_hierarchical_2020}.

@@

\subsection{Removal Modelling}
\par Estimates of cue rate changed very little from species that were originally modelled by the single-species models in \citet{edwards_point_2023}, to the same ones modelled in the multi-species model here.
The 5-fold cross validation tended to show no preference for one model or another when sample size was large, and a preference (albeit an extremely small preference) for single-species models when sample size was smaller.
When sample size was small, a multi-species model tended to produce more precise estimates of cue rate.
Given the extremely small preference for single-species models under a lower sample size, it would appear that the bias-variance trade-off is likely a welcome trade-off here, because the decrease in variance is met by only a very slight increase in bias.

\par There were seven species where the relative difference in cue rate between the single-species model and multi-species model was over 20\% (noting that this threshold was chosen arbitrarily).
These species (Willow Ptarmagin, Rusty Blackbird, Red-breasted Sapsucker, Common Redpoll, Calliope Hummingbird, Allen's Hummingbird, and Bohemian Waxwing), mostly had fairly small sample sizes ($\leq 515$ for all species except Red-breasted Sapsucker [$n = 1854$] and Common Redpoll [$n = 2009$]), had the lowest cue rate in the single-species models, and had among the lowest cue rates in the multi-species model.
For removal modelling, all those species except for Common Redpoll ended up having the multi-species model chosen as the more predictive model, indicating that sharing information among related species tended to improve their estimates of cue rate.
Common Redpoll was likely an exception to this because the original estimate of cue rate in NA-POPS was a wild underestimate (cue rate of $5.9\times 10^{-9}$), likely because of a convergence issue in the \texttt{detect} package.
It could be worth running Common Redpoll using the Bayesian single species model developed here to see if the estimate of cue rate is still extremely low, or if the vague but still somewhat informative prior helps to bring the estimated parameter into a more sensible space.
It could also be worth investigating the merit of this type of modelling to estimate cue rate for species where the cue rate is extremely low.
In our case, 41 out of the 79 species that had the lowest 25\% quantile of cue rate in the single-species model ended up having the multi-species model chosen as the preferred model.

\par Although this was not a primary goal of our study, we found much less support than \citet{solymos_phylogeny_2018} for migration strategy having an effect on cue rate.
Both the estimate effects for resident species and migratory species were effectively 0, with very little of the density above or below 0.025 from 0.
One potential reason is that we had proportionally more resident species in our analysis (85 out of 316; 26.9\%) compared to \citet{solymos_phylogeny_2018} (23 out of 141; 16.3\%), and so the richer information provided by these additional species could have brought the effect down to 0.
Additionally, while \citet{solymos_phylogeny_2018} only regressed estimates of cue rate against the predictor variables, we actually estimated cue rate within the context of the same predictors, and so the phylogenetic effect could have just overwhelmed the estimates, making the migratory strategy meaningless.

\subsection{Distance Modelling}
\par Estimates of EDR also changed very little, albeit a non-zero amount, from species that were originally modelled by the single-species models in \citet{edwards_point_2023}, to the same ones modelled in the multi-species model here.
However, there was evidence for an average small downward shift in EDRs when modelled under a multi-species context.
Still, the 5-fold cross validation tended to show almost no preference for one model or another when the sample size was large, and a slight (but still very small) preference for the multi-species model when the sample size was small.
There was less evidence for the multi-species model being able to improve the precision of EDR estimates across sample sizes, indicating that a single-species model may mostly be sufficient for precise estimates.
Under this context, it would appear that a multi-species model may slightly bring down estimates of EDR for species with a smaller sample size.

\par There were six species where the relative difference in EDR between the single-species model and multi-species model was over 20\%. 
Unlike removal modelling, these species (Sharp-tailed Grouse, Sedge Wren, Ring-necked Pheasant, Chimney Swift, Boat-tailed Grackle, and Baird's Sparrow) had little in common with each other in terms of sample size (range of 227 for Boat-tailed Grackle through 4466 for Baird's Sparrow) and EDR (ranging from 50.8 m to 226 m in single-species, and from 73.2 m to 159.1 m in the multi-species model). 
Additionally, there was mixed support among these species for single-species vs. multi-species: Chimney Swift and Boat-tailed Grackle had the multi-species model preferred, whereas the others had the single-species model preferred.
It could be noted that both Chimney Swift and Boat-tailed Grackle showed an increase in estimated EDR from single-species to multi-species modelling and preferred the multi-species model, but Baird's Sparrow also showed an increase and preferred the single-species model.

\par Similar to removal modelling, we found less support than \citet{solymos_phylogeny_2018} for migration strategy having an effect on EDR, although the posterior distribution for both the resident and migrant species were much further away from zero, indicating at least some support.
Habitat type, body mass, and song pitch, as expected, all had an overall effect on EDR, as was found in \citet{solymos_phylogeny_2018} and \cite{johnston_species_2014}.

\subsection{Relaxation of Sample Size}

\par For five out of the seven species where we generated new predictions of detectability in this study, we were interested in seeing how a multi-species QPAD removal model could estimate cue rate with few data, and how a multi-species QPAD distance model could estimate EDR with few data.

\par Overall, the estimates of cue rate fell roughly where we would expect based on related species, for species where we were simply relaxing the sample size requirements.
For example, Harris's Sparrow (n = 65) had an estimated cue rate of 0.30 cues per minute, with a 95\% credible interval of [0.21, 0.42], compared to the overall mean cue rate of 0.37 cues per minute for all 32 other Passerellidae species in NA-POPS.
Similarly, LeConte's Thrasher (n = 1) had an estimated cue rate of 0.32 cues per minute, with a 95\% credible interval of [0.18, 0.57].
Though this was a much larger interval than Harris's Sparrow, it still falls within a reasonable distance from the overall mean cue rate of 0.36 cues per minute for all 6 other Mimidae species in NA-POPS.

>>>>

\par To test our prediction that there would be little change between estimates of species already modelled in NA-POPS, we ran the Bayesian hierarchical QPAD Removal and Distance models on the set of species for which there was sufficient data in the original NA-POPS analysis.
In this case, we ran both models for 316 species of birds.
Rufous Hummingbird (\textit{Selasphorus rufus}) was dropped from the distance modelling due to issues with divergent transitions \citep{betancourt_diagnosing_2016, leimkuhler_simulating_2005}.
Tricoloured Blackbird had sufficient distance data, and so was included in the distance modelling, but not the removal modelling.
We accessed the single-species estimates of cue rate and EDR from the NA-POPS Null models using the \texttt{napops} R package \citep{edwards_napops_2023, edwards_point_2023}.
We then modelled the difference in estimated cue rates and EDRs as
$$\Delta X \sim N(0,\sigma)$$
$$\sigma \sim exp(1)$$

where $\Delta X$ is calculated as the mean multi-species estimate of cue rate or EDR minus the mean single-species estimate of cue rate of EDR.
If our prediction is true, we would expect to see a distribution of modelled differences centred about 0, indicating no real difference in estimates of cue rate or EDR between single-species and multi-species models, for species for which we have sufficient data.

\par To test our prediction about improved precision, we simply plotted the species-specific estimates of standard deviations of cue rate and EDR, respectively, against the log sample size, for each of the single-species and multi-species model. 
We then fit a smooth LOESS curve through each of the model-specific points.
If our prediction is true, we would expect to see a plot showing a higher standard deviation in single-species models than for multi-species models, but only up until a certain sample size when they should both be similar.

\par To test our prediction about improved accuracy, we ran a k-fold cross-validation with 5 folds, stratified by species.
That is, for each cross-validation run, we held out 20\% of the data for each species for both the multi-species and Bayesian single-species models. 
We then derived multi-species estimates of cue rate and EDR and single-species estimates of cue rate and EDR, for each cross-validation run, and used these parameter estimates to calculate the estimated pointwise log posterior density of each of the held out data points. 
We then calculated the difference in posterior predictive density for removal and distance models, by subtracting the pointwise posterior predictive density of the single-species model from the pointwise posterior predictive density of the multi-species model (i.e., multi-species minus single-species).
Finally, we took the mean and standard error of the difference in pointwise log posterior density for each species to create a species-specific mean and standard error of difference in pointwise posterior density.
With this, positive intervals (i.e., where mean $\pm$ standard error is all positive) would indicate that the multi-species model was preferred, and negative intervals would indicate that the single-species model was preferred.
If our prediction is true, then we would expect to see majority positive values of the species-specific mean pointwise posterior density.

\subsection{Predicting Detection Probabilities}
\par To predict estimates of cue rate or EDR from the multi-species models, we adopted two possible approaches.
If the NA-POPS database contained at least some point-count data for the species of interest, we simply added these data into the model.
This is different from \citet{edwards_point_2023, solymos_calibrating_2013, solymos_evaluating_2018}, where any species with less than 75 data points were removed from the analysis.
For species that did not have any data in the NA-POPS database, we generated a row of traits data for the species and estimated its cue rate and EDR as parameters in the model.
We then ran the multi-species models described above using these additional data points (and hence an updated phylogenetic correlation structure for the removal model) to generate estimates of cue rate and EDR for these new species.

\subsection{Analysis}
\par All analyses were performed in R version 4.2.2 \citep{r_core_team_r_2022}.
Manipulation of phylogenetic trees, including finding a consensus tree for visualization and generating variance-covariance matrices based on the 2000 bootstrapped trees, was done using the \texttt{ape} R package \citep{paradis_ape_2019}.
All Bayesian models were coded in the Stan Probabilistic Language \cite{stan_development_team_stan_2019}, and models were run using the \texttt{cmdstanr} R package \citep{gabry_cmdstanr_2022} and visualized using the \texttt{bayesplot} R package \citep{gabry_visualization_2019}.
Each of the single-species models and multi-species models, as well as the models created to assess performance, were run using 1000 warmup iterations and 2000 sampling iterations, each on 4 chains for a total of 8000 draws per model run.
To mitigate issues surrounding initial values being rejected in the distance model (see Discussion), we set the initial values of EDR to be the mean EDR from NA-POPS estimates; for species where EDR estimates did not exist in NA-POPS, we used the overall mean EDR as initial value.
Cross-validation models were run with 1000 warmup iterations and 500 sampling iterations, each on 4 chains for a total of 2000 draws per model run.
All of the models made use of Stan's reduce\_sum functionality, which allows for within-chain parallelization as well as the usual across-chain parallelization.
This was particularly useful in our case as the multi-species removal model contained over 3 million data points and the multi-species distance model contained over 4 million data points, and so computational expense was a large concern.
All models were run on one of two multi-core processing servers, one running Ubuntu 20.04.4 LTS and one running Ubuntu 20.04.6 LTS.

@@

\par We found some evidence that a multi-species removal model improved the precision of cue rate estimates compared to estimates from a single-species model (Figure \ref{fig:sd}A).
This was particularly evident for species with sample sizes less than approximately $\exp\left\{7\right\} \approx 1097$, where the estimated LOESS curve for single-species standard deviation was always higher than the estimated LOESS curve for multi-species standard deviations.
For sample sizes greater than 1097, standard deviations appeared similar in both the single-species and multi-species model, indicating no real improvement in precision.

\par We found little evidence that a multi-species model improves the precision of EDR compared to that of a single-species model (Figure \ref{fig:sd}B).
Both estimated LOESS curves for single-species standard deviations and multi-species standard deviations were roughly equal across all sample sizes, indicating no real improvement (or decrease) in precision.

@@

\par The cross validation revealed that multi-species model had better predictive capacity for more species (200 out of 316) than the single-species model (99 out of 316).
Seventeen out of 316 species showed no better predictive capacity with either model.
Mean overall difference in log predictive density was $4.78 \times 10^{-6}$.
As expected, multi-species models tended to have better predictive capacity when sample size was low, with a negligible difference as sample size increased (Figure \ref{fig:cv}A).
There did not appear to be any patterns in model selection among phylogeny or traits (see Supplemental Figure 1 for full phylogenetic tree).

\par (NEEDS UPDATING) The cross validation revealed that the multi-species model tended to have slightly more support than the single-species model in terms of predicting EDR.
However, similar to the removal model cross-validation results, the overall difference in posterior predictive density was very small, with a mean difference of $1.85 \times 10^{-4}$.
Multi-species models tended to have better predictive capacity when sample size was low (again noting a very small difference), and a negligible difference as sample size increased (Figure \ref{fig:cv}B).
Interestingly, only two (Chimney Swift and Boat-tailed Grackle) of the six species with a greater than 20\% change in EDR had the multi-species model chosen as the better predictive model.

@@

\par The multi-species removal model produced reasonable estimates of cue rates for all seven species considered here (Figure \ref{fig:predictions}A).
Species where the sample size requirements were relaxed (i.e., all except Bicknell's Thrush and Kirtland's Warbler) had reasonable precision associated with the estimates.
One exception to this was LeConte's Thrasher, where the precision was rather low.
The species for which we had no data (i.e., Bicknell's Thrush and Kirtland's Warbler) had estimates of cue rate similar to that of their closely-related species, demonstrating the shrinkage effect of the multi-species model.

\par The multi-species distance model also produced reasonable estimates of EDR for all six species considered here (Figure \ref{fig:predictions}B).
Species where the sample size requirements were relaxed (i.e., all except Bicknell's Thrush) had fairly precise estimates, except for LeConte's Thrasher and Lesser Prairie Chicken where the estimated EDR spanned greater than 50 m in either direction.
The species for which we had no data (i.e., Bicknell's Thrush) had an estimate of EDR similar to what would be expected by a bird of its size, song pitch, and habitat association, demonstrating the shrinkage effect of the multi-species model.

@@

\par In both cases of QPAD removal modelling and QPAD distance modelling, we were able to generate estimates of detectability for species with no data, and precise estimates of detectability for data-sparse species.
Overall, the shrinkage effect offered by the multi-species modelling approach allowed for more accurate estimates of cue rate and EDR to be derived in species with smaller sample sizes, without affected estimates in species with large sample sizes.
The best model for predictive accuracy tended to depend on the species, but overall differences in the estimates were small enough that using a multi-species model should not produce extremely inaccurate estimates.

\par Although this was not a primary goal of our study, when considering the species-level covariates, we found much less support than \citet{solymos_phylogeny_2018} for migration strategy having an effect on cue rate and EDR.
Both the estimated effects for resident species and migratory species were effectively 0 in the removal model, with very little of the density above or below 0.025 from 0.
These estimates varied slightly more in the distance model, but the 95\% credible interval still captured 0.
One potential reason is that we had proportionally more resident species in our analysis (85 out of 316; 26.9\%) compared to \citet{solymos_phylogeny_2018} (23 out of 141; 16.3\%), and so the richer information provided by these additional species could have brought the effect down to 0.
Habitat type, body mass, and song pitch, as expected, all had an overall effect on EDR, as was found in \citet{solymos_phylogeny_2018} and \cite{johnston_species_2014}.

\subsection{Predicting Detectability for Data-sparse Species}

\par For five out of the seven species for which we generated new predictions of detectability in this study, we were interested in seeing how a multi-species QPAD removal model could estimate cue rate with few data, and how a multi-species QPAD distance model could estimate EDR with few data.

\par Overall, the estimates of cue rate fell roughly where we would expect based on related species, for species for which we were simply relaxing the sample size requirements.
For example, Harris's Sparrow (n = 65) had an estimated cue rate of 0.30 cues per minute, with a 95\% credible interval of [0.21, 0.42], compared to the overall mean cue rate of 0.37 cues per minute for all 32 other Passerellidae species in NA-POPS.
Similarly, LeConte's Thrasher (n = 1) had an estimated cue rate of 0.32 cues per minute, with a 95\% credible interval of [0.18, 0.57].
Though this was a much larger interval than Harris's Sparrow, it still falls within a reasonable distance from the overall mean cue rate of 0.36 cues per minute for all 6 other Mimidae species in NA-POPS.

\par Estimates of EDR also fell roughly where we would expect based on species of similar traits, for the species for which we were relaxing samples sizes.
For example, Harris's Sparrow (n = 13) had an estimated mean EDR of 65.6 m, with a 95\% credible interval of [53.1, 82.5].
This range does fall below the mean EDR of 85.3 m for the other sparrows in NA-POPS, but Harris's Sparrow song pitch of approximately 3.8 Hz falls above the average song pitch of 3.0 Hz for the other sparrows in NA-POPS, which could contribute to the smaller EDR.
On the other hand, LeConte's Thrasher (n = 1) had an estimated EDR of 134.2 m, with a 95\% credible interval of [81.8, 230.5], which falls roughly in line with the mean EDR of 113.2 m of the six other thrashers in NA-POPS.

\par It is likely that the width of the credible intervals (i.e., the precision) of these relaxed sample-size species will depend on the number of data points available for that particular species, the number of "similar" species, and the number of data points available for those similar species.
Additionally, it is important to note that we did not account for other sources of variation at the survey level, such as when the survey was conducted (i.e., time of day, time of year), or where the survey was conducted (e.g., in a forest, by a roadside).
It is well known that the ability to detect a species can be very dependent on the time of day or year that an observer is surveying, as well as the environmental conditions of the survey \citep{edwards_point_2023}.
Thus, some of the variation in these estimates of cue rate and EDR could be due to not accounting for survey-level variation that affects detectability, and so it is important that future iterations of this model incorporate both species-level effects on detectability \textit{and} survey-level effects on detectability.

\par We generated predictions of cue rate and EDR for both Bicknell’s Thrush and Kirtland’s Warbler.
For both species, we had zero removal sampling data, and so cue rates were solely based on information sharing among phylogeny (and to a much lesser extent, migration status). 
For Kirtland’s Warbler, we did have one data point for distance sampling, so predictions of EDR for Kirtland’s Warbler here are not entirely based on zero data points.
Bicknell’s Thrush, however, had zero data points, and so EDRs for this species were based solely on information sharing among other species with similar traits.

\par As expected, the predicted cue rate for Bicknell’s Thrush (0.28 cues per minute) and the predicted cue rate for Kirtland’s Warbler (0.29 cues per minute) fell very close to the overall mean cue rate of 0.27 for all 11 other Turdidae species and 0.31 for all 44 other Parulidae species in NA-POPS, respectively. 
That is, in the absence of any data, the estimated cue rate of Bicknell’s Thrush and Kirtland’s Warbler was shrunk toward cue rates of most similar species.

\par For distance modelling, the predicted EDR for Bicknell's Thrush (73.6 m) and the predicted EDR for Kirtland's Warbler (75.3 m) fell reasonably close to the overall mean EDR of 92.7 m for all other 11 Turdidae species and 61.2 m for all 44 other Parulidae species in NA-POPS, respectively.
hat is, in the absence of any data (or with very little data in the case of Kirtland’s Warbler), the estimated EDR of Bicknell’s Thrush and Kirtland’s Warbler was similar to those of other Thrushes and Warblers, respectively, but with departures from overall mean based on specific traits.

\par In both the removal modelling and distance modelling, the magnitude of contribution of data from related species appeared to play a role in the width of the credible intervals. 
For removal modelling, although both credible intervals were fairly wide, Bicknell’s Thrush had a 95\% credible interval of [0.16, 0.48], representing an approximate 53\% change on either end, whereas Kirtland's Warbler had a 95\% credible interval of [0.18, 0.48], representing an approximate 49\% change on either end.
For distance modelling, both credible intervals were also fairly wide, but Bicknell's Thrush had a 95\% credible interval of [40.5, 134.4], representing an approximate 58\% change on either end, whereas Kirtland's Warbler had a 95\% credible interval of [45.9, 125.6], representing an approximate 47\% change on either end.
Although the difference in width of the intervals is small, this could be due to the greater number of other species contributing information to Kirtland's Warbler estimates than to Bicknell's Thrush.
In the case of distance modelling, the one data point contributing to Kirtland's Warbler's EDR could have also played a part.
With that in mind, it would likely be important to consider how many "related" species are available with information (and how many data points do each of those related species have) when deciding on making a prediction of cue rate.

\par Using the predictions of cue rate and EDR for Bicknell’s Thrush and Kirtland’s Warbler, we can calculate availability and perceptibility, and hence detectability, for a given survey type, and compare to previous estimates of detectability obtained from other studies.
As a reminder, following derivations in \citet{solymos_calibrating_2013}, detectability is the product of availability and perceptibility.
Availability $p(t)$ is defined as the probability of a bird giving a cue during a survey run for $t$ minutes, and is calculated as
$p(t) = 1 - e^{-t\phi}$
where $\phi$ is the cue rate of the species.
Perceptibility $q(r)$ is defined as the probability of an observer hearing or seeing a bird, given the bird gives a cue, with a survey radius of $r$ metres, and is calculated as
$q(r) = \dfrac{\tau^2 \left\{1 - e^{-\dfrac{r^2}{\tau^2}}\right\}}{r^2}$
where $\tau$ is the EDR of the species.

\par For Bicknell's Thrush, using the equations above, the predicted cue rate of 0.28, predicted EDR of 73.3 m, and an example survey design of 5 minutes recording all birds within 100 m, we estimate a detectability of 0.34.
We can also use the lowest predicted cue rate and EDR and the highest predicted cue rate and EDR (based on the respectively credible intervals) to get a range of detectabilities from 0.09 through 0.70.
\citet{aubry_bicknells_2018}, however, estimate detection probabilities of $\geq$ 0.74 and $\leq$ 0.88, and so even our highest potential value of detectability falls short of their estimated range.
However, we note that \citet{aubry_bicknells_2018} are using data that specifically target Bicknell's Thrush in their range, and they are using surveys conducted at peak Bicknell's Thrush breeding time, and so it is likely that detectability will be high in that specific case.

\par For Kirtland's Warbler, using the equations above, the predicted cue rate of 0.29, predicted EDR of 75.3 m, and an example survey design of 5 minutes recording all birds within 100 m, we estimate a detectability of 0.36.
Similar to our Bicknell's Thrush example, we can also use the lowest predicted cue rate and EDR and the highest predicted cue rate and EDR (based on the respective credible intervals) to obtain a range of detectabilities from 0.12 through 0.67.
A study by \citet{van_dyke_comparative_2022} found estimated detection probabilities for Kirtland’s Warbler that ranged from effectively 0, through to a detection probability of 0.73
In that study, they compared estimates of detectability among different jack pine stands and red pine stands, where the detectability was higher in jack pine stands than the red pine stands.

@@

\par The multi-species QPAD Distance model had several issues related to initialization, sampling, and convergence throughout this study.
For many iterations of the model development, generation of initial values using Stan proved difficult.
For all unconstrained parameters, Stan draws initial values from a uniform(-2,2) distribution.
On the scale that we modelled detection distances (i.e., log detection distance in hundreds of metres), these are reasonable values.
At the low end, an initial value of -2 would correspond to a detection distance of $e^{-2} \times 100 = 13.5$ m, which is low but somewhat plausible for a small-bodied, high-pitched species such as a hummingbird.
At the high end, an initial value of 2 would correspond to a detection distance of $e^{2} \times 100 = 739$ m, which is high but still somewhat plausible for large-bodied, low-pitched species like a grouse or prairie-chickens.
However, where Stan seemed to struggle is when these randomly assigned values for detection distances were effectively impossible for a large set of species.
That is to say, if the values were initialized such that hummingbirds had a detection distance of 500 m or more, and grouse or prairie-chickens had a detection distance of 15 m or less, this would result in an impossible likelihood, forcing Stan to reject the initial values.
Our fix was to set initial values to be jittered mean detection distances originally estimated in \citet{edwards_point_2023}; for species without estimates, we used the mean EDR for all species.
This approach seemed to work as the initial values were no longer rejected, but we caution that this could very well be a "bandaid solution" to an underlying issue with the model itself.

\par The distance models also seemed to struggle with sampling the posterior distribution, in that over a quarter of the iterations resulted in a divergent transition \citep{betancourt_diagnosing_2016, leimkuhler_simulating_2005}.
However, after several iterations of this model, we identified that the samplers were not appropriately sampling Rufous Hummingbird's EDR, in that there appeared to be a self-imposed barrier in being able to sample EDRs less than 25 m. 
Once we removed Rufous Hummingbird from the analysis, all divergent transitions vanished.
We demonstrate some of the odd sampling of the posterior for Rufous Hummingbird in Supplemental Materials B.
One potential explanation is that Rufous Hummingbird had the smallest EDR of all the species in NA-POPS, and so there could be an issue with numerical stability of sampling very low values, due to the exponentiation of a quotient in the distance modelling likelihood.
Future studies should seek to test the limits of this distance model, particularly in a Bayesian setting, using simulated data with very low EDRs.

@@
